---
title: "BS730 R Markdown Example"
author: "Chathurangi Pathiravasan"
date: "August 22, 2020"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

```{r libraries, include=FALSE}
library(knitr)
```


## Analysis of Iris Petal and Sepal Size

In this report, we will assess the correlation between **petal** and **sepal** size for different species of iris.
It will produce the following: 

* Summary statisitcs for each variable in the iris dataset
* Correlations
    + Between sepal length and petal length 
    + Between sepal width and petal width
* Plot of sepal length vs. petal length 

***

### Summary Statistics

#### Without kable function
```{r summary}
data(iris)
summary(iris)
```

#### With kable function
```{r summary_kable}
data(iris)
kable(summary(iris))
```

There are `r nrow(iris)` flowers in the dataset.

***

### Correlation between sepal and petal size

```{r correlations}
lcorr <- cor(iris$Sepal.Length, iris$Petal.Length)
lcorr
wcorr <- cor(iris$Sepal.Width, iris$Petal.Width)
wcorr
```

The correlation between sepal length and petal length is `r round(lcorr, 2)`  
The correlation between sepal width and petal width is `r round(wcorr, 2)`

***

### Plot of petal length vs. sepal length

```{r length_plot, fig.height=5, fig.width=7, fig.path='Figures/', echo=FALSE}
plot(iris$Sepal.Length, iris$Petal.Length,
     xlab = "Sepal Length", ylab = "Petal Length")
```

